Equivalence Testing and the Second Generation \emph{P}-Value

### 2018-08-28 manuscript.tex
submission

 % Line numbering
  \usepackage{lineno}
  \linenumbers

@@

  \abstract{To move beyond the limitations of null-hypothesis tests, statistical
approaches have been developed where the observed data is compared
against a range of values that are equivalent to the absence of a
meaningful effect. Specifying a range of values around zero allows
researchers to statistically reject the presence of effects large enough
to matter, and prevents practically insignificant effects from being
interepreted as a statistically significant difference. We compare the
behavior of the recently proposed second generation \emph{p}-value
(Blume, D'Agostino McGowan, Dupont, \& Greevy, 2018) with the more
established Two One-Sided Tests (TOST) equivalence testing procedure
(Schuirmann, 1987). We show that the two approaches yield almost
identical results under optimal conditions. Under suboptimal conditions
(e.g., when the confidence interval is wider than the equivalence range,
or when confidence intervals are asymmetric) the second generation
\emph{p}-value becomes difficult to interpret as a descriptive
statistic. The second generation \emph{p}-value is interpretable in a
dichotomous manner (i.e., when the SGPV equals 0 or 1 because the
confidence intervals lies completely within or outside of the
equivalence range), but this dichotomous interpretation does not require
calculations. We conclude that equivalence tests yield more consistent
\emph{p}-values, distinguish between datasets that yield the same second
generation \emph{p}-value, and allow for easier control of Type I and
Type II error rates.}
  \keywords{equivalence testing, second generation \emph{p}-values, hypothesis
testing, TOST, statistical inference \\

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV6-1.pdf}
\caption{\label{fig:TOSTSGPV6}Comparison of \emph{p}-values from TOST (black
line) and SGPV (dotted grey line) across a range of observed sample
means (x-axis). Because the sample size is small (n = 10) and the CI is
more than twice as wide as the equivalence range (set to -0.4 to 0.4),
the SGPV is set to 0.5 (horizontal lightgrey line) across a range of
observed means.}
\end{figure}

@@

The effect of asymmetric confidence intervals is most noticable at
smaller sample sizes. In Figure \ref{fig:TOSTSGPV11} we plot the
\emph{p}-values from equivalence tests and the SGPV (again plotted as
1-SGPV for ease of comparison) for correlations. The sample size is 30
pairs of observations, and the lower and upper equivalence bounds are
set to -0.45 and 0.45, with an alpha of 0.05. As the observed
correlation in the sample moves from -1 to 0 the \emph{p}-value from the
equivalence test becomes smaller, as does 1-SGPV. The pattern is quite
similar to that in Figure \ref{fig:TOSTSGPV2}. The \emph{p}-value for
the TOST procedure and 1-SGPV are still identical when \emph{p}-values
are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted
lines). There are two important differences, however. First of all, the
SGPV is no longer a straight line, but a curve, due to the asymmetry in
the 95\% CI. Second, and most importantly, the \emph{p}-value for the
equivalence test and the SGPV do no longer overlap at \emph{p} = 0.5.

>>>>

@@

  \abstract{To move beyond the limitations of null-hypothesis tests, statistical
approaches have been developed where the observed data is compared
against a range of values that are equivalent to the absence of a
meaningful effect. Specifying a range of values around zero allows
researchers to statistically reject the presence of effects large enough
to matter, and prevents practically insignificant effects from being
interpreted as a statistically significant difference. We compare the
behavior of the recently proposed second generation \emph{p}-value
(Blume, D'Agostino McGowan, Dupont, \& Greevy, 2018) with the more
established Two One-Sided Tests (TOST) equivalence testing procedure
(Schuirmann, 1987). We show that the two approaches yield almost
identical results under optimal conditions. Under suboptimal conditions
(e.g., when the confidence interval is wider than the equivalence range,
or when confidence intervals are asymmetric) the second generation
\emph{p}-value becomes difficult to interpret as a descriptive
statistic. The second generation \emph{p}-value is interpretable in a
dichotomous manner (i.e., when the SGPV equals 0 or 1 because the
confidence intervals lies completely within or outside of the
equivalence range), but this dichotomous interpretation does not require
calculations. We conclude that equivalence tests yield more consistent
\emph{p}-values, distinguish between datasets that yield the same second
generation \emph{p}-value, and allow for easier control of Type I and
Type II error rates.}
  \keywords{equivalence testing, second generation \emph{p}-values, hypothesis
testing, TOST, statistical inference \\

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/TOSTSGPV6-1.pdf}
\caption{\label{fig:TOSTSGPV6}Comparison of \emph{p}-values from TOST (black
line) and SGPV (dotted grey line) across a range of observed sample
means (x-axis). Because the sample size is small (n = 10) and the CI is
more than twice as wide as the equivalence range (set to -0.4 to 0.4),
the SGPV is set to 0.5 (horizontal light grey line) across a range of
observed means.}
\end{figure}

@@

The effect of asymmetric confidence intervals is most noticeable at
smaller sample sizes. In Figure \ref{fig:TOSTSGPV11} we plot the
\emph{p}-values from equivalence tests and the SGPV (again plotted as
1-SGPV for ease of comparison) for correlations. The sample size is 30
pairs of observations, and the lower and upper equivalence bounds are
set to -0.45 and 0.45, with an alpha of 0.05. As the observed
correlation in the sample moves from -1 to 0 the \emph{p}-value from the
equivalence test becomes smaller, as does 1-SGPV. The pattern is quite
similar to that in Figure \ref{fig:TOSTSGPV2}. The \emph{p}-value for
the TOST procedure and 1-SGPV are still identical when \emph{p}-values
are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted
lines). There are two important differences, however. First of all, the
SGPV is no longer a straight line, but a curve, due to the asymmetry in
the 95\% CI. Second, and most importantly, the \emph{p}-value for the
equivalence test and the SGPV do no longer overlap at \emph{p} = 0.5.
